---
title: "Day 3: R programming"
author: "Aaron R. Williams (IBP)"
date: "`r format(Sys.time(), '%B %d, %Y %H:%M')`"
output:
  html_document:
    number_sections: TRUE
    self_contained: TRUE    
    code_folding: show
    toc: TRUE
    toc_float: TRUE
    css: ../www/web_report.css
    editor_options:
      chunk_output_type: console
---

<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>

<link href="https://fonts.googleapis.com/css?family=Lato&display=swap" rel="stylesheet">

# Review

* `select()`
* `filter()`
* `rename()`
* `arrange()`
* `mutate()`
* `summarize()`
* `group_by()`
* `%>%`

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Motivation

* Review and master `library(dplyr)`
* Expand our skill set so we can do more types of analysis

### Exercise 0

<font color="#55b748">**Step 1:**</font> Open up your `.Rproj` for `urbn101`. You should now see `urbn101` in the top right of RStudio instead of “Project: (None)” . 

<font color="#55b748">**Step 2:**</font> Install the R package `nycflights13`. 

<font color="#55b748">**Step 3:**</font> Open a `.R` script and save it as `03_flights-analysis.R`.

<font color="#55b748">**Step 4:**</font> Load `tidyverse` and `nycflights13` at the top of your script. 

<font color="#55b748">**Step 5:**</font> Add and run `data(flights)` to load the flights data set. Add and run `data(planes)` to load the planes data set. 

<font color="#55b748">**Step 6:**</font> Use `glimpse()` and `View()` to look at the `flights` data set.

# `mutate()`

Last week, we showed how `mutate()` can be used to create new variables or to transform new variables. For example, we converted knots into miles per hour. In this section, we will walk through some useful transformations on the `flights` data set.  

## Conditional transformation

`case_when()` allows for a sequence of conditional logic that can be used to transform or create variables with `mutate()`. For example, here I create a variable called `lateness` that creates a character variable with the levels `"very late"`, `"late"`, and `"on time"`. The logic is evaluated from top to bottom and `TRUE` can be used as otherwise. The logical test goes to the left of `~` and the result goes to the right. `&` and `|` can be used to combine logical statements. 

```{r eval = FALSE}
flights %>%
  mutate(
    lateness = case_when(
      arr_delay > 30 ~ "very late",
      arr_delay > 0 ~ "late",
      TRUE          ~ "on time"
    )
  )

```

### Exercise 1

`dep_delay` is the departure delay in minutes and `arr_delay` is the arrival delay in minutes. 

<font color="#55b748">**Step 1:**</font> Set up `mutate()` and `case_when()` after a pipe from the `flights` data set. We are going to create a variable called `delay_status`. 

<font color="#55b748">**Step 2:**</font> Create a condition for if `dep_delay` and `arr_delay` are both `> 0` (use `&` to combine) and assign the value `"both late"`. Use `count()` to count the number of observations with `"both late"`. There should be `NA`. 

<font color="#55b748">**Step 3:**</font> Create a condition for if `dep_delay` and `arr_delay` are both `< 0` (use `&` to combine) and assign the value `"both early"`.

<font color="#55b748">**Step 4:**</font> Create an otherwise with `TRUE` and assign the value `"mixed"`. 

<font color="#55b748">**Step 5:**</font> Use `count()` to count the number of flights with each level of `delay_status`. 

## Strings

`library(stringr)` contains powerful and concise functions manipulating character variables. Reference the cheat sheet for a survey of the different possible transformations. 

### Exercise 2

`month` and `day` are currently integer variables. We want to turn them into character variables with leading zeros. For example, `1` should be `"01"`.

<font color="#55b748">**Step 1:**</font> Use `mutate()` to overwrite `month` and `day` with `str_pad()`. The first argument should be `month` or `day`. The second argument, `width`, should be `2`. 

<font color="#55b748">**Step 2:**</font> The padding character is currently a space, but we want it to be `"0"`. Use `?str_pad` to figure out how to switch the padding character. 

<font color="#55b748">**Step 3:**</font> Pipe the result into the following line `mutate(flight_date = paste(year, month, day, sep = "-"))`

<font color="#55b748">**Step 4:**</font> Drop all variables except `flight_date`, `distance`, and `air_time`. 

<font color="#55b748">**Step 5:**</font> Assign the result to `flights_subset`. 

## Dates

`library(lubridate)` contains powerful and concise functions for creating and manipulating dates, times, and date-times. It is aware of leap days and leap seconds and is useful for calculating periods, durations, intervals, and more. 

### Exercise 3

<font color="#55b748">**Step 1:**</font> Add `library(lubridate)` after `library(nycflight13)` in your script. 

<font color="#55b748">**Step 2:**</font> `library(lubridate)` is powerful but it needs variable in the correct format. Use `ymd()` inside of `mutate()` to create a new variable called `flight_date` for the `flights_subset` data. 

<font color="#55b748">**Step 3:**</font> Use `mutate()` and `wday(flight_date)` to find the day of the week for each date.

<font color="#55b748">**Step 4:**</font> Assign the result to `flights_subset`. 

<font color="#55b748">**Step 5:**</font> Use `count()` to count the number of flights by day of the week. 

# `summarize()`

### Exercise 4

We are going to summarize `flights_subset` from the previous example. `summairze()` can be used to create multiple statistics for one variable and statistics for many variables at the same time. 

<font color="#55b748">**Step 1:**</font> `group_by()` `week_day` and use `n()` in `summarize()` to count the number of observations. This should match Step 5 from the previous exercise. 

<font color="#55b748">**Step 2:**</font> In the same `summarize()`, calculate `mean()`, `median()`, and `max()` `distance`.

<font color="#55b748">**Step 3:**</font> In the same `summarize()`, calculate mean and median `air_time`.

<font color="#55b748">**Step 4:**</font> Rename the resulting variables *inside* `summarize()` so they have more useful names. 

### Survey data

`library(srvyr)`, a tidy extension of Thomas Lumley's `library(survey)`, allows for complex survey designs to be tabulated with `summarize()`. [This guide](https://ui-research.github.io/urbn101-intro-r/lessons/08_survey-analysis) contains background and a few examples. 

# `left_join()`

Joins are the main method for combining data sets. There are many types of joins. We will focus on the "left" join, which merges observations from the right data set to the left data set. Let's look at the cheat sheet.  

```{r}
students <- tribble(
  ~name, ~subject, 
  "Aaron", "Math",
  "Kyle", "History",
  "Ajjit", "Math",
  "Sarah", "Science"
)

teachers <- tribble(
  ~subject, ~teacher,
  "Math", "Mr. Smith",
  "History", "Ms. Johnson",
  "Science", "Mr. Robot"
)

left_join(x = students, y = teachers, by = "subject")

```

### Exercise 5

`flights` contains information about flights and the unit of observation is airplane flights. `planes` contains information about the airplanes and the unit of observation is the airplane. Common units of observation at Urban are household, person, state, and county.

We want to add information about planes to the flights data set. We will perform a many-to-one join because we are joining planes multiple times to each unique flight. The above example is also a many-to-one join. 

<font color="#55b748">**Step 1:**</font> Use `left_join()` to join `planes` to `flights`. The common key is `tailnum`. 

<font color="#55b748">**Step 2:**</font> Use `left_join()` to see observations from `flights` that don't have a match in `planes`. The common key is `tailnum`. 

<font color="#55b748">**Step 3:**</font> Use `filter()` to see if a random `tailnum` from the `anti_join()` is in `planes`. Hint: it shouldn't be!

We highly recommend reading the [chapter on joins in R4DS](https://r4ds.had.co.nz/relational-data.html). It will inform all statistical programming. 

# Custom functions

Sometimes functions don't exist for desired calculations or we want to combine many calculations into one function to reduce copying-and-pasting. 

"You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)." ~ [R4DS](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

R has a flexible function system.

```
function_name <- function(arg1, arg2 = default) {
  # function body
}
```

Three ingredients

* Function name - usually verbs
* Function arguments - inputs to the function (optional)
* Function body

### Example

```{r}
square <- function(x = 2) {
  x ^ 2
}

square()

square(x = 4)
```

**Note:** Use tidyverse functions inside of custom functions often requires non-standard evaluation. Please reach out for help when this is your goal. 

### Exercise 7

<font color="#55b748">**Step 1:**</font> Write a function called `multiply_xy()` that takes arguments `x` and `y` and multiplies them together. 

<font color="#55b748">**Step 2:**</font> Add your favorite number as the default for `x` and your least favorite number as the default for `y`.

<font color="#55b748">**Step 3:**</font> Call the function and overwrite the default for `y` with your favorite number.


# System functions

R contains a full set of functions for managing the computer system. 

* `file.create()`
* `file.exists()`
* `file.remove()`
* `file.rename()`
* `file.copy()`
* `dir.create()`

### Example

```
if (!file.exists("data/cps.csv")) {
  download.file("https//:cps-download.gov", "data/cps.csv")
}
```

### Exercise 8

<font color="#55b748">**Step 1:**</font> Check to see if you are in an R project. Use `file.exists()` to confirm that your `.R` script exists.  

<font color="#55b748">**Step 2:**</font> Use `file.rename()` to rename your R script. 

<font color="#55b748">**Step 3:**</font> Use `file.copy()` to duplicate your script and then delete the copy with `file.remove()`. CAREFUL! There is no undo button!

# Conditional logic

Conditional logic is an important concept in computer programming. We already used `ifelse()` and `case_when()` to create indicator variables and conditional variables. Sometimes it's also useful to run entire chunks of code conditionally.

```
if () {

} else if () {

} else {

}
```

Three ingredients

* Conditional predicate (`if`, `else if`, `else`)
* Conditional statement (`==`, `>`, `<`, `%in%`)
* Code body

### Example

```{r}
x <- 1000

if (x > 999) {
  "x is big"
} else {
  "x is small"
}
```

### Example

```{r}
urbnmapr::get_urbn_map
```

### Exercise 9

<font color="#55b748">**Step 1:**</font> Create a custom function called `is_it_me` with one argument called `name`.

<font color="#55b748">**Step 2:**</font> Create a conditional control structure with `if` and `else`. 

<font color="#55b748">**Step 3:**</font> If `name ==` your name then return "It's me!". If `name != ` your name then return "It's not me!". 

# Friday: data merges

# Resources

* [R4DS: functions](https://r4ds.had.co.nz/functions.html)
