---
title: "Day 4: R programming"
author: "Aaron R. Williams (IBP)"
date: "`r format(Sys.time(), '%B %d, %Y %H:%M')`"
output:
  html_document:
    number_sections: TRUE
    self_contained: TRUE
    code_folding: show
    toc: TRUE
    toc_float: TRUE
    css: ../www/web_report.css
    editor_options:
      chunk_output_type: console
---

<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>

<link href="https://fonts.googleapis.com/css?family=Lato&display=swap" rel="stylesheet">

# Review

* What is `<-` and why do we use it?
* What is tidy data?
* How do we create an indicator variable with `mutate()`?

```{r message = FALSE}
library(tidyverse)
```

# Motivation

* R contains a rich set of flexible tools for progamming.
* Most programs in R are the combination of relatively simple concepts.
* Survey of building blocks for R programming.

# Math operations

## Vectors

Until now, we've mostly focused on data frames. Functions like `filter()` and `mutate()` all take data frames as arguments and can accept vectors in arguments without re-referencing their data frame.

```{r}
msleep %>%
  mutate(name = tolower(name))
```

Base R and some R tasks require vectors instead of data frames. We'll focus on two main methods of pulling vectors frome data frames and a method for creating a vector from scratch.

```{r}
# $ operator
msleep$sleep_total

# pull()
msleep %>%
  pull(sleep_total)

# review: build your own vector
c(8, 10, 1000, pi)
```

## Math functions

R contains many functions for common mathematical operations. 

* `sum()` = $\sum_{i = 1}^n x_i$
* `prod()` = $\Pi_{i = 1}^n x_i$
* `mean()`
* `var()`
* `sd()`
* `cov()`
* `cor()`
* `median()`
* `min()`
* `max()`

Math functions in R expect vectors. We can use them in `summarize()` without the above syntax.

```{r}
summarize(msleep, max(sleep_total))
```

We can also use math functions with the following syntax.

```{r}
mean(msleep$sleep_total)

msleep %>%
  pull(sleep_total) %>%
  median()

min(c(8, 10, 1000, pi))
```

<font color="#55b748">**Step 1:**</font> Open up a script, save it with a meaningful name, and load the tidyverse.

<font color="#55b748">**Step 2:**</font> Find the mean of `sleep_rem` from the `msleep` data set.

<font color="#55b748">**Step 3:**</font> Find the 25th, 37th, and 92nd percentiles of `awake` from `msleep` with one function call.

# Custom functions

Sometimes functions don't exist for desired calculations or we want to combine many calculations into one function to reduce copying-and-pasting. 

"You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)." ~ [R4DS](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

R has a flexible function system.

```
function_name <- function(arg1, arg2 = default) {
  # function body
}
```

Three ingredients

* Function name - usually verbs
* Function arguments - inputs to the function (optional)
* Function body

```{r}
square <- function(x = 2) {
  x ^ 2
}

square()

square(x = 4)
```

**Note:** Use tidyverse functions inside of custom functions often requires non-standard evaluation. Please reach out for help when thsi is your goal. 

<font color="#55b748">**Step 1:**</font> Write a function called `multiply_xy()` that takes arguments `x` and `y` and multiplies them together. 

<font color="#55b748">**Step 2:**</font> Add your favorite number as the default for `x` and your least favorite number as the default for `y`.

<font color="#55b748">**Step 3:**</font> Call the function and overwrite the default for `y` with your favorite number.

# String manipulation

Character variables contain strings that are important for many programing tasks like person names, managing hundreds of `.csv` files, or analyzing text.  

We often want to perform operations on these strings.

## paste0()

```{r}
paste("apples", "and", "oranges")

directory <- "data/"
file_name <- "criminal-justice.csv"
paste0(directory, file_name)

```

<font color="#55b748">**Step 1:**</font> Combine your first name, middle initial, and last name into one text string.

## library(stringr)

`paste()` and `paste0()` are two of many functions for string manipulation. 

`library(stringr)` is a nearly comprehensive set of functions and tools for string manipulation. 



## Vectorized

String functions are mostly vectorized. So iterating a function on each object is painless. 

```{r}
names <- c("Kyle", "Sarah", "Aaron")

paste0("data/", names, ".csv")
```

# System functions

R contains a full set of functions for managing the computer system. 

* `file.create()`
* `file.exists()`
* `file.remove()`
* `file.rename()`
* `file.copy()`
* `dir.create()`

```
if (!file.exists("data/cps.csv")) {
  download.file("https//:cps-download.gov", "data/cps.csv")
}
```
<font color="#55b748">**Step 1:**</font> Check to see if you are in an R project. Use `file.exists()` to confirm that your `.R` script exists.  

<font color="#55b748">**Step 2:**</font> Use `file.rename()` to rename your R script. 

<font color="#55b748">**Step 3:**</font> Use `file.copy()` to duplicate your script and then delete the copy with `file.remove()`. CAREFUL! There is no undo button!

# Conditional logic

Conditional logic is a foundational concept in computer programming. We already used `ifelse()` and `case_when()` to create indicator variables and conditional variables. Sometimes it's also useful to run entire chunks of code conditionally.

```
if () {

} else if () {

} else {

}
```

Three ingredients

* Conditional predicate (`if`, `else if`, `else`)
* Conditional statement (`==`, `>`, `<`, `%in%`)
* Code body

```{r}
urbnmapr::get_urbn_map
```

# Friday: data merges

# Resources


