---
title: "Day 4: R programming"
author: "Aaron R. Williams (IBP)"
date: "`r format(Sys.time(), '%B %d, %Y %H:%M')`"
output:
  html_document:
    number_sections: TRUE
    self_contained: TRUE
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
    css: ../www/web_report.css
    editor_options:
      chunk_output_type: console
---

<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>

<link href="https://fonts.googleapis.com/css?family=Lato&display=swap" rel="stylesheet">

# Review



# Motivation


# Math functions

## Vectors



```{r}
library(tidyverse)

msleep$sleep_total

msleep %>%
  pull(sleep_total)
```

## Functions

```{r}
mean(msleep$sleep_total)

quantile(msleep$sleep_total, prob = c(0.1, 0.2))
```

<font color="#55b748">**Step 1:**</font> Open up a script, save it with a meaningful name, and load the tidyverse.

<font color="#55b748">**Step 2:**</font> Find the mean of `sleep_rem` from the `msleep` data set.

<font color="#55b748">**Step 3:**</font> Find the 25th, 37th, and 92nd percentiles of `awake` from `msleep` with one function call.

# Custom functions

"You should consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)." ~ R4DS


```
function_name <- function(arg1, arg2 = default) {

}
```

# String manipulation


# System commands


# Conditional logic

# Friday: data merges



# Resources


